---
title: 'Analyzing Statistical Arbitrage Methods: A Comparative Study (Code)'
author: "Lorenzo Spinazzi"
date: "2024-05-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Packages used:
```{r, warning=FALSE, message = FALSE}
library(quantmod)
library(lubridate)
library(readr)
library(urca)
library(ggplot2)
library(reshape2)
library(tseries)
library(e1071)
library(openxlsx)
```

# Data

## SPY Data Download: 
```{r, message=FALSE, warning=FALSE}
# Download adjusted closing prices of SPY members
# Downloading SPY tickers from CSV file
tickers <- read_csv("SPY Tickers.csv")
tics <- c()
for(i in 1:nrow(tickers)){
  tics <- append(tics, paste(tickers[i,1]))
}

# Getting adjusted closing price data from Yahoo Finance
P_list <- lapply(tics, function(x) getSymbols(x, from = "2010-01-01", auto.assign = FALSE)[,6])
P <- Reduce(merge, P_list)
colnames(P) <- tics

# Eliminate stocks with NA values
elim_list <- c()
for(i in 1:ncol(P)){
  if(is.na(P[1,i]) == TRUE){
    elim_list <- c(elim_list, i)
  }
}
P <- P[, -elim_list]

# Calculate (simple) returns
R <- (P - lag(P)) / lag(P)
R <- R[-1,]
```

## Random Date Selection Function:
```{r}
# Inputs:
# "R" <- xts object of return time series
date.selection <- function(R){
  # Ensure after starting formation day there is at least 1.5 years of data
  R2 <- R["2010-01-01/2022-01-01"]
  n <- nrow(R2)
  # Random selection of starting formation day
  start <- sample(n, 1)
  # Formation start and end dates
  formation.start <- index(R[start,])
  formation.end <- seq(formation.start, by = "year", length.out = 2)[2]
  # Trading start and end dates
  trade.start <- formation.end + 1
  trade.end <- trade.start %m+% months(6)
  # Formation interval (1 year)
  interval.1 <- paste(formation.start, "/", formation.end, sep = "")
  # Trading interval (6 months)
  interval.2 <- paste(trade.start, "/", trade.end, sep = "")
  # Formation + Trading interval
  interval.3 <- paste(formation.start, "/", trade.end, sep = "")
  intervals <- c(interval.1, interval.2, interval.3)
  return(intervals)
}
```

# Distance Method

## Price Path Normalization Function
```{r}
# Inputs:
# "R" <- xts object of return time series
# "intervals" <- vector of intervals outputted from date.selection function
normalize <- function(R, intervals){
  R.cum <- cumprod(1 + R[intervals[3], ]) - 1
  return(R.cum)
}
```

## SSD Pairs Selection Function
```{r}
# Inputs:
# "R" <- normalized return series FORMATION PERIOD ONLY
# N <- desired number of pairs (top N lowest SSD)
pairs.selection <- function(R, N){
  min.list <- c(rep(Inf, N))
  pairs.list <- data.frame()
  for(i in 1:(ncol(R) - 1)){
    for(j in (i + 1):(ncol(R))){
      P1.name <- colnames(R[,i])
      P2.name <- colnames(R[,j])
      SSD.hat <- as.numeric(sum((R[,i] - R[,j])^2))
      spread <- R[,i] - R[,j]
      spread.mu <- mean(spread)
      spread.sigma <- sd(spread)
      if(nrow(pairs.list) < N){
        pairs.list <- rbind(pairs.list, c(P1.name, P2.name, SSD.hat, spread.mu, spread.sigma))
        min.list[which.max(min.list)] <- SSD.hat
      }
      else if(SSD.hat < max(min.list)){
        pairs.list[which.max(pairs.list[,3]), ] <- c(P1.name, P2.name, SSD.hat, spread.mu, spread.sigma)
        min.list[which.max(min.list)] <- SSD.hat
      }
    }
  }
  colnames(pairs.list) <- c("Stock 1", "Stock 2", "SSD", "Mean Spread", "Spread SD")
  return(pairs.list)
}
```

## Trading Function
```{r}
# Inputs:
# "pairs" <- outputted from pairs.selection
# "R.cum.trade" <- normalized returns TRADING PERIOD ONLY
# "R.trade" <- regular returns (not normalized) TRADING PERIOD ONLY
# "plots" <- boolean indicating whether or not you want outputted plots (T = yes)
trade.strat <- function(pairs, R.cum.trade, R.trade, plots){
  
  # Initializing data frame of pair position returns
  P.returns <- data.frame(rep(0, nrow(R.trade)))
  P.returns <- as.xts(P.returns, order.by = index(R.trade))
  
  # Calculating the return time series for every pair in the portfolio
  for(i in 1:nrow(pairs)){
    
    # Returns of stock 1 during trading period
    s1.r <- as.numeric(R.trade[, which(colnames(R.trade) == pairs[i,1])])
    
    # Normalized (cumulative returns) returns of stock 1 during trading period
    s1.cr <- as.numeric(R.cum.trade[, which(colnames(R.cum.trade) == pairs[i,1])])
    
    # Returns of stock 2 during trading period
    s2.r <- as.numeric(R.trade[, which(colnames(R.trade) == pairs[i,2])])
    
    # Normalized (cumulative returns) returns of stock 2 during trading period
    s2.cr <- as.numeric(R.cum.trade[, which(colnames(R.cum.trade) == pairs[i,2])])
    
    # Normalized spread
    spread <- s1.cr - s2.cr
    
    # Mean spread during formation period
    mu <- as.numeric(pairs[i, 4])
    
    # Spread standard deviation during formation period
    sig <- as.numeric(pairs[i, 5])
    
    # Initializing trading signals (begin with neutral signal)
    signal.list <- c(0)
    signal <- 0 
    
    # Initializing vector of returns from long-short positions
    return <- c()
    
    # Initialize which stock we are longing/shorting (this doesn't matter)
    long.s <- s1.cr
    short.s <- s2.cr

    for(j in 1:nrow(R.trade)){

################################################################################     
######### If spread goes above two historical standard deviations ##############
################################################################################
      if((spread[j] >= (mu + 2 * sig)) & signal == 0){
        
        # Record the trading signal
        signal.list <- c(signal.list, -1)
        signal <- -1
        
        # No return at this time step given positions are entered the following
        # trading day
        return[j] <- 0
        
        # Determination of long stock and short stock
        # Long "loser" short "winner"
        if(s1.cr[j] > s2.cr[j]){
          long.s <- s2.cr
          long.r <- s2.r
          short.s <- s1.cr
          short.r <- s1.r
        } else {
          long.s <- s1.cr
          long.r <- s1.r
          short.s <- s2.cr
          short.r <- s2.r
        }
      } 
################################################################################     
### If we are currently in an open position (-1) and paths haven't converged ###
################################################################################
      else if(signal == -1 & (spread[j] > mu)){
        
        # Record the trading signal (remains -1)
        signal.list <- c(signal.list, -1)
        signal <- -1
        
        # Calculate daily return
        return[j] <- long.r[j] - (short.r[j]) 
      }

################################################################################      
#### If we are currently in an open position (-1) and paths have converged #####
################################################################################
      else if(signal == -1 & (spread[j] <= mu)){
        
        # Record trading signal (back to neutral 0 for next day)
        signal.list <- c(signal.list, 0)
        signal <- 0
        
        # Calculate daily return
        return[j] <- long.r[j] - (short.r[j])
      }
################################################################################      
######### If spread goes below two historical standard deviations ##############
################################################################################
      else if((spread[j] <= (mu - 2 * sig)) & signal == 0){
        signal.list <- c(signal.list, 1)
        signal <- 1
        return[j] <- 0
        if(s1.cr[j] > s2.cr[j]){
          long.s <- s2.cr
          long.r <- s2.r
          short.s <- s1.cr
          short.r <- s1.r
        } else {
          long.s <- s1.cr
          long.r <- s1.r
          short.s <- s2.cr
          short.r <- s2.r
        }
      } 
################################################################################     
### If we are currently in an open position (1) and paths haven't converged ####
################################################################################ 
      else if(signal == 1 & (spread[j] < mu)){
        signal.list <- c(signal.list, 1)
        signal <- 1
        return[j] <- long.r[j] - (short.r[j])
      }
################################################################################      
#### If we are currently in an open position (1) and paths have converged ######
################################################################################
      else if(signal == 1 & (spread[j] >= mu)){
        signal.list <- c(signal.list, 0)
        signal <- 0
        return[j] <- long.r[j] - (short.r[j])
      }
################################################################################
#### We are in a neutral position and the spread hasn't triggered a trade ######     
################################################################################
      else{
        signal.list <- c(signal.list, 0)
        signal <- 0
        return[j] <- 0
      }
    }

################################################################################
############ If we want graphical representations of the process ###############  
################################################################################    
    if(plots == TRUE){
      
      # Plot normalized spread 
      plot(spread, type = "l", main = paste(pairs[i,1], "/", pairs[i,2], " Spread ",  sep = ""), ylab = "Spread")
      abline(h = mu, col = "black", lty = 2)
      abline(h = (mu + 2 * sig), col = "red", lty = 2)
      abline(h = (mu - 2 * sig), col = "red", lty = 2)
      
      plot(s1.cr, type = "l", col = "blue", main = paste(pairs[i,1], "/", pairs[i,2], " Normalized Price Paths",  sep = ""), ylab = "Normalized Price", ylim = c(min(c(s1.cr, s2.cr)), max(c(s1.cr, s2.cr))))
      lines(s2.cr)
      
      # Plot trade signal evolution
      plot(signal.list, type = "l", main = paste(pairs[i,1], "/", pairs[i,2], " Trade Signals ",  sep = ""), ylab = "Signal")

    }
    
    # Add return time series to portfolio of returns
    P.returns <- cbind(P.returns, return)
  }
  
  # Return portfolio of returns
  P.returns <- P.returns[,-1]
  return(P.returns)
}
```

## Pairs Portfolio Return Function
```{r}
# Inputs:
# "P.returns" <- output from trade.strat function
portfolio.return <- function(P.returns){
  N <- ncol(P.returns)
  w0 <- rep(1/N, N)
  w <- data.frame()
  w <- rbind(w, w0)
  r <- as.numeric(P.returns[1,])
  P.r <- c()
  P.r[1] <- sum(w0 * r)
  for(i in 2:nrow(P.returns)){
    r <- as.numeric(P.returns[i - 1, ])
    ri <- as.numeric(P.returns[i,])
    wi <- as.numeric(w[i - 1, ] * (1 + r))
    w <- rbind(w, wi)
    P.r[i] <- as.numeric(sum(wi * ri))
  }
  return(P.r)
}
```

# Cointegration Method

## Correlated Pairs Selection Function
```{r}
# Inputs:
# "R.cum.form" <- normalized returns FORMATION PERIOD ONLY
correlation_pairs <- function(R.cum.form){
  Sigma <- cor(R.cum.form)
  cor_pairs <- data.frame()
  for(i in 1:(nrow(Sigma)-1)){
    for(j in (i+1):ncol(Sigma)){
        if(Sigma[i,j] >= 0.80){
          asset1 <- rownames(Sigma)[i]
          asset2 <- colnames(Sigma)[j]
          cor_pairs <- rbind(cor_pairs, c(asset1, asset2))
      }
    }
  }
  colnames(cor_pairs) <- c("Asset 1", "Asset 2")
  return(cor_pairs)
}
```

## Cointegrated Pairs Selection Function
```{r, warning=FALSE}
# Inputs:
# "cor_pairs" <- output from correlation_pairs function
# "R.cum.form" <- normalized returns FORMATION PERIOD ONLY
cointegration <- function(cor_pairs, R.cum.form){
  coin_pairs <- data.frame()
  for(i in 1:nrow(cor_pairs)){
    asset1.name <- cor_pairs[i,1]
    asset2.name <- cor_pairs[i,2]
    asset1 <- R.cum.form[, which(colnames(R.cum.form) == cor_pairs[i,1])]
    asset2 <- R.cum.form[, which(colnames(R.cum.form) == cor_pairs[i,2])]
    
    # Cointegration test
    test <- summary(ca.jo(data.frame(asset1, asset2), type = "trace", K = 2, ecdet = "none", spec = "longrun"))
    if((is.na(test@teststat[2]) == FALSE) & (is.na(test@cval[2,3]) == FALSE)){
      if(test@teststat[2] > test@cval[2,3]){
        row.entry <- c(asset1.name, asset2.name, test@teststat[2], test@cval[2,3], test@lambda[1], test@V[1,1], test@V[2,1])
        coin_pairs <- rbind(coin_pairs, row.entry)
      }
    }
  }
  colnames(coin_pairs) <- c("Asset 1", "Asset 2", "Test Stat", "1 pct CL", "Lambda", "Asset 1 Eigen", "Asset 2 Eigen")
  coin_pairs <- coin_pairs[order(coin_pairs[, 3], decreasing = TRUE), ]
  return(coin_pairs)
}
```

## Stationary Pairs Selection Function
```{r, warning=FALSE}
# Inputs:
# "coin_pairs" <- output from cointegration function
# "R.cum.form" <- normalized returns FORMATION PERIOD ONLY
# "N" <- top N most significant results (from Johansen Test)

stationarity <- function(coin_pairs, R.cum.form, N){
  stationary_pairs <- data.frame()
  pair_num <- 0
  i <- 0
  
  while(pair_num < N){
    i <- i + 1
    asset1.name <- coin_pairs[i,1]
    asset2.name <- coin_pairs[i,2]
    asset1 <- R.cum.form[, which(colnames(R.cum.form) == coin_pairs[i,1])]
    asset2 <- R.cum.form[, which(colnames(R.cum.form) == coin_pairs[i,2])]
    
    spread <- as.numeric(asset1 - asset2)
    SSD <- NA
    spread.mu <- mean(spread)
    spread.sig <- sd(spread)
    
    asset1 <- asset1 * as.numeric(coin_pairs[i,6])
    asset2 <- asset2 * as.numeric(coin_pairs[i,7])
    pair_path <- asset1 + asset2
    
    test <- adf.test(pair_path)
    
    if(test$p.value == 0.01){
      row.entry <- c(asset1.name, asset2.name, SSD, spread.mu, spread.sig)
      stationary_pairs <- rbind(stationary_pairs, row.entry)
      pair_num <- pair_num + 1
    }
  }
  colnames(stationary_pairs) <- c("Asset 1", "Asset 2", "SSD", "Average Spread", "Spread SD")
  return(stationary_pairs)
}
```

# Ornstein Uhlenbeck Method

## Ornstein-Uhlenbeck Method Parameter Tune Function
```{r}
# Inputs:
# "X" <- alpha/beta scaled portfolio time series values
# "dt" <- time increment value (i.e., daily = 1/252)
OU_tune <- function(X, dt){
  # Leung and Li (2015) analytical derivations
  N <- length(X)
  Xx <- sum(X[1:(N-1)])
  Xy <- sum(X[2:N])
  Xxx <- sum((X[1:(N-1)])^2)
  Xxy <- sum(X[1:(N-1)] * X[2:N])
  Xyy <- sum(X[2:N]^2)

  # Optimal theta, mu, and sigma parameters 
  theta <- (Xy * Xxx - Xx * Xxy) / (N * (Xxx - Xxy) - (Xx^2 - Xx * Xy))

  mu <- (-1 / dt) * log((Xxy - theta * Xx - theta * Xy + N * theta^2) / (Xxx - 2 * theta * Xx + N * theta^2))

  sigma <- sqrt((2 * mu / (N * (1-exp(-2*mu*dt)))) * (Xyy - 2* exp(-mu * dt) * Xxy + exp(-2 * mu * dt) * Xxx - 2 * theta * (1-exp(-mu*dt)) * (Xy - Xx * exp(-mu*dt)) + N * theta^2 * (1-exp(-mu * dt))^2))

  tune <- c(theta, mu, sigma)
  return(tune)
}
```

## Log-Likelihood Function
```{r}
# Inputs:
# "X" <- alpha/beta scaled portfolio time series values
# "tune" <- output from OU_tune
# "dt" <- time increment value (i.e., daily = 1/252)
log.likelihood <- function(X, tune, dt){
  theta <- tune[1]
  mu <- tune[2]
  sigma <- tune[3]
  N <- length(X)
  sig.tilde <- sqrt(sigma^2 * ((1 - exp(-2 * mu * dt)) / (2 * mu)))
  
  f_sum <- 0
  for(i in 2:N){
    x2 <- X[i]
    x1 <- X[i - 1]
    
    f_sum <- f_sum + (x2 - x1 * exp(-mu * dt) - theta * (1 - exp(-mu * dt)))^2
  }
  
  output <- -0.5 * log(2 * pi) -log(sig.tilde) - (1 / (2 * N * sig.tilde^2)) * f_sum
  return(output)
}
```

## Optimal Beta Value Function
```{r}
# Inputs:
# "s1" <- price time series of long position stock
# "s2" <- price time series of short position stock
# "dt" <- time increment value (i.e., daily = 1/252)
alpha.beta.optim <- function(s1, s2, dt){
  s1 <- as.numeric(s1)
  s2 <- as.numeric(s2)
  ratio <- seq(0.001, 1, 0.001)
  f <- c()
  a <- 1 / s1[1]
  for(i in 1:length(ratio)){
    b <- ratio[i] / s2[1]
    X <- a * s1 - b * s2
    tune <- OU_tune(X, dt)
    f[i] <- log.likelihood(X, tune, dt)
  }
  max <- ratio[which(f == max(f))]
  b <- max / s2[1]
  alpha.beta <- c(a,b)
  return(alpha.beta)
}
```

## Ornstein Trading Function
```{r}
# Inputs
# "pairs" <- selected pair
# "intervals" <- vector of intervals outputted from date.selection function 
# "dt" <- time increment value (i.e., daily = 1/252)
# "plots" <- boolean indicating whether or not you want outputted plots (T = yes)
OU.trade <- function(pairs, intervals, dt, plots){

  # Stock pair chosen from SSD criteria
  asset1 <- P[, which(colnames(P) == pairs[1,1])]
  asset2 <- P[, which(colnames(P) == pairs[1,2])] 
  pair <- cbind(asset1, asset2)
  
  # Formation period prices
  s1.form <- as.numeric(pair[intervals[1],1])
  s2.form <- as.numeric(pair[intervals[1],2])
  
  # OU parameter calculation
  alpha.beta <- alpha.beta.optim(s1.form, s2.form, dt)
  alpha <- alpha.beta[1]
  beta <- alpha.beta[2]
  
  # Forming OU process
  OU.form <- alpha * s1.form - beta * s2.form
  
  # Trade signals
  mu <- mean(OU.form)
  sigma <- sd(OU.form)
  
  # Trading period prices
  s1.trade <- as.numeric(pair[intervals[2],1])
  s2.trade <- as.numeric(pair[intervals[2],2])
  
  # OU process for trading period
  OU.test <- alpha * s1.trade - beta * s2.trade
 
  # Initializing signal and return vectors 
  signal <- 0
  signal_list <- c(0)
  returns <- c()
  
  for(i in 2:length(OU.test)){
    
    # OU process hits bottom threshold
    if((OU.test[i] <= (mu - 2 * sigma)) & (signal == 0)){
      # Long spread signal initiated for next iteration
      signal <- 1
      signal_list[i] <- 1
      returns[i - 1] <- 0
    }
    
    # Long spread active, remain in long spread (calculate return)
    else if((OU.test[i] < mu) & (signal == 1)){
      signal <- 1
      signal_list[i] <- 1
      returns[i - 1] <- ((s1.trade[i] - s1.trade[i - 1]) / s1.trade[i - 1]) + ((s2.trade[i - 1] - s2.trade[i]) / s2.trade[i])
    }
    
    # Long spread active, mean reversion, close long spread position, still calculate this day's return
    else if((OU.test[i] >= mu) & (signal == 1)){
      signal <- 0
      signal_list[i] <- 0
      returns[i - 1] <- ((s1.trade[i] - s1.trade[i - 1]) / s1.trade[i - 1]) + ((s2.trade[i - 1] - s2.trade[i]) / s2.trade[i])
    }
    
    # OU process hits top threshold
    else if((OU.test[i] >= (mu + 2 * sigma)) & (signal == 0)){
      # Short spread signal initiated for next iteration
      signal <- -1
      signal_list[i] <- -1
      returns[i - 1] <- 0     
    }
    
    # Short spread active, remain in short spread (calculate return)
    else if((OU.test[i] > mu) & (signal == -1)){
      signal <- -1
      signal_list[i] <- -1
      returns[i - 1] <- ((s2.trade[i] - s2.trade[i - 1]) / s2.trade[i - 1]) + ((s1.trade[i - 1] - s1.trade[i]) / s1.trade[i])
    }
    
    # Short spread active, mean reversion, close short spread position, still calculate this day's return
    else if((OU.test[i] <= mu) & (signal == -1)){
      signal <- 0
      signal_list[i] <- 0
      returns[i - 1] <- ((s2.trade[i] - s2.trade[i - 1]) / s2.trade[i - 1]) + ((s1.trade[i - 1] - s1.trade[i]) / s1.trade[i])
    }
    
    # No signal triggered and we have no active positions
    else if(signal == 0){
      signal <- 0
      signal_list[i] <- 0
      returns[i - 1] <- 0
    }
    
  }
  
  # Plot results 
  if(plots == TRUE){
    OU.overall <- c(OU.form, OU.test)
    plot(OU.overall, type = "l", main = paste(colnames(asset1), " + ", colnames(asset2), " Ornstein-Uhlenbeck Trade", sep = ""), ylab = "OU Process", xlab = "Day")
    abline(h = mu, col = "red", lty = 2)
    abline(h = (mu + sigma), col = "blue", lty = 2)
    abline(h = (mu - sigma), col = "blue", lty = 2)
    abline(v = length(s1.form), lty = 2, col = "green")
    
    legend("topleft", legend = c("OU Process", "Trade Signals", "Long Term Average", "Formation/Trading Split"), col = c("black", "blue", "red", "green"), lty = c(1,2,2,2), cex = 0.8)
    
    plot(signal_list, type = "l", ylab = "Signal", xlab = "Day")
    abline(h = 1, lty = 2, col = "red")
    abline(h = -1, lty = 2, col = "green")
    abline(h = 0, lty = 2, col = "black")
  }
    return(returns)
}
```


